 

### 2021-01-06 vertical-thermal-review.Rmd
attempting to organize paragraphs

Forests are vertically and horizontally stratified, with trees of different sizes through various successional stages. Overstory trees form canopies that play a crucial role in moderating macro-microclimate [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017] by buffering extreme temperatures along with other macroclimatic conditions in the understory. This creates a vertical stratification of biophysical environments such as temperature, light, wind, humidity and Co2 which influences leaf temperature and metabolism along the gradient.[Bertrand et al. 2011]

@@

*(paragraph on leaf area distribution, including figure ([separate from or combined with current Fig. 2- see GitHub issue #15](https://github.com/EcoClimLab/vertical-thermal-review/issues/15))*
The vertical structure and hence, microenvironments of forest canopies vary seasonally (Parker & Tibbs 2004; Smith et al. 2019 ; Parker et al. 2019) and following disturbance (e.g., Parker et al. 2002 ; Almeida et al. 2016 ; Stark et al. 2020 in press).
In this review, we focus primarily on growing season conditions, including leaf phenology at the edges of the growing season but excluding periods (in deciduous forests) when no leaves are present.

Light conditions, specifically photosynthetically active radiation (PAR) varies along the vertical gradient with leaf area index (LAI), canopy height, canopy structure, across species and forest types due to cumulative layers of vegetation [@koike_leaf_2001]. Generally, upper canopies in dense forests are exposed to greater PAR values and limit light to canopy interiors, lower canopy layers and understories. Shaded regions under canopy cover experience lower values and quality of PAR [in ranges of blue, orange and red]. The gradient of light is more pronounced under broadleaved forests than conifers[@aussenacInteractionsForestStands2000a; @niinemetsPackingPhotosyntheticMachinery2009; Parker 1995; @poorterMetaanalysisPlantResponses2019b]
(Fig. `r fig_NEON_vertical`).
In heterogeneous canopies [with high gap fractions], distance from the outer canopy is a better proxy for light environment than height (Parker 1995) and lower canopy light environments are highly variable, ranging from high light gap to deeply shaded understory regions [@chazdonPhotosyntheticLightEnvironments1984; @tymenQuantifyingMicroenvironmentalVariation2017; @smithSeasonalDroughtrelatedChanges2019].

Tropical forests with dense overstory limit light to understories and subsequent layers, where tropical forest floors receive approximately 1-2%  of the light incident to the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], whereas mixed-wood boreal and Northern temperate forest floors receive around 4-5% of full sunlight [@bartemucciOverstoryInfluencesLight2006].
This difference in light regime along a vertical profile is an important driver of plasticity in leaf traits, leaf physiology, adaptation, and thermal sensitivity along the gradient. Leaves along a vertical intra-canopy gradient thus exhibit a variation in leaf trait adapaptions to a range of high light to shaded-ness [@michaletzEnergeticCarbonEconomic2016; @fauset_differences_2018; @niinemetsResponsesForestTrees2010, @sackHowStrongIntracanopy2006; @doughtyAreTropicalForests2008; Niinemets and Valladares 2006].

@@

Wind speed in the upper canopy of a semi-deciduous tropical forest has been observed to be higher in dry season than in upper canopy wet season [@rey-sanchez_spatial_2016] [*double check*]. 
Greater upper canopy wind speeds result in higher boundary layer conductance ($g_{a}$) for canopy leaves, which increases linearly with increase in wind velocity [@daudetWindSpeedLeaf1999a]. Large $g_{a}$ to heat transfer results in smaller difference in $T_{air}$ and $T_{leaf}$, this is more typical of small leaves in upper canopy that make them more advantageous in dissapating heat efficiently than larger leaves[@bauerleSeparatingFoliarPhysiology2011]. In low wind conditions, because $g_{a}$ to heat transfer is small enough, $T_{leaf}$ in upper canopy can become substantially elevated than $T_{air}$ under high radiation [@martin_boundary_1999]. Additionally, thickness of air layer is proportional to boundary layer resistance at the surface of the leaf through which water vapor diffuses after leaving the stomata [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a; @stokesBoundaryLayerConductance2006].

@@

Similarly, maximum air temperatures were higher above than below tropical forest canopies during wet seasons in Panama [@rey-sanchez_spatial_2016], temperate forests in Switzerland [@vonarxSpatiotemporalEffectsForest2012a] and coastal Brazil [@fauset_differences_2018, Tymen et al. 2017].

"max temps were higher and humidity lower in gaps vs understory regions (both at 1.5 m above the ground), see Fig. 4 - again illustrating buffering effect of the canopy".However, similar maximum temperatures have been observed during the dry season of a semi-deciduous tropical forest in Panama partially because in the dry season some canopy trees were leafless, resulting in more light reaching the sub-canopy, and more air movement between subcanopy and above-canopy air[@rey-sanchez_spatial_2016], similar results were observed in in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate].

@@

With rising temperatures, this trend in maximum $T_{air}$ and RH would have implications for understory seedling growth, especially in high altitude temperate pine forests and montane forests  [@vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018] (to clarify why? are these high elevation forests limited by low temperature and low rh?)

RH, in combination with temperature, determines vapor pressure deficit, $VPD$. Atlantic forests during wet season tended to have higher $T_{air}$, and $VPD$ in upper canopy and canopy gaps than lower in understory where $T_{air}$ and $VPD$ were lower [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017].  
This means that canopy leaves tend to be exposed to higher *evaporative demand*, higher stomtal contrains and lower osmotic potential. [@niinemetsPhotosyntheticAcclimationSimultaneous2004a]. In North Eastern US temperate forests, upper canopy was observed to be more sensitive to $VPD$, whereas as understory was insensitive [@rollinsonClimateSensitivityUnderstory, 2020]

**Finally, carbon dioxide (CO~2~) concentrations tend to be higher in the understory.** 
-	Higher in understory, particularly at dusk (Koike et al. 2001).
-	Higher in the understory overnight; difference persists during the day but is very small [@yang_spatial_1999].
- Differences in concentration are by far most pronounced near ground level, as well, concentrations of (CO~2~) mole fraction values have been observed to be higher in lower canopy leaves compared with upper canopy leaves [@niinemetsPhotosyntheticAcclimationSimultaneous2004a]

@@

*Presumably, buffering would also be affected by canopy roughness, which affects turbulent air flow and the canopy boundary layer.*
Taller trees don't necessarily increase buffering [@zellweger_seasonal_2019]. 
Species that have higher shade casting ability such as braodleaved trees increase T buffering [@zellweger_seasonal_2019,@vonarxSpatiotemporalEffectsForest2012]
The strength of buffering also varies with respect to geographic and climatic factors.
* Distance to coast, topographic position, elevation [@zellweger_seasonal_2019]
* [@davis_microclimatic_2019]--Proximity to riparian areas tends to sustain buffering, protecting understory temperatures even during increasing temperatures [@davis_microclimatic_2019]
However vertical climate gradients within forests are known to be steeper than those driven by elevation and latitude [@nakamuraForestsTheirCanopies2017; @martinCovarianceSunShade2020a]

@@

Leaf traits are shaped by intra-canopy light and height, and show intra-canopy foliar plasticity in accordance with the intensity and availability of light. Degrees of high irradiance gives rise to sun-leaf morphology compared to degrees of low irradiance that is characterized by shade shapes shade-leaf morphology [@poorterMetaanalysisPlantResponses2019b; Sack et al 2006] within- and across species [Ishii et al. 2018; @niinemetsWorldwideAnalysisWithincanopy2015]. The degree of intra-canopy plasticity also varies across species and geographically [@sackHowStrongIntracanopy2006; @ishiiVariationIntracrownLeaf2018].

Cateogorization of sun and shade tolerance from @niinemetsControversyTraitsConferring2006 are as follows: "The five-level scale used for shade tolerance (1,very intolerant; 2, intolerant; 3, moderately tolerant; 4,tolerant; 5, very tolerant) corresponds approximately to the following light availabilities expressed as percentage of full sunlight: 1,>50%; 2, 25-50%; 3, 10-25%; 4, 5-10%; 5, 2-5%". Additionally, shade leaves in the understory are of three categories; 1. shade leaves of canopy trees--top of the tree is sun exposed, bottom is shaded. 2. seedlings/saplings of canopy trees that will eventually, if they survive, reach the canopy. 3 understory plants that complete their lifecycle in the shaded understory. These groups are likely to have different traits, and if we expect 'shade' plants to have a different sensitivity to climate change than sun leaves, the consequences of climate change for demography and species community composition of the forest will depend strongly on the category of shade plants being considered.[refs, -Martijn Slot]

@@

Taller trees in dense forests tend to be exposed to greater irradiance than lower canopy and shorter understory trees shaded by overstory.As a result the different combination of stressors taller trees with foliage exposed to greater PAR are different from the combination of stressors experienced by understory shorter trees with shaded/semi-shaded leaves [Niinemetes et al. 2010] 
"Spending their lifecycle in the shade means they have different pressures [such as competition for light or adapting leaf traits to maximum light capture and use efficiency] than plants that get the bulk of their carbon from sun-exposed leaves, but can maintain a few layers of shade leaves as well."[-Martijn]
Additionally, increased exposure to light increases carbon reserve, plant height, foliage robustness,  while also increasing evapotranspiration, decreasing hydraulic conductivity, increasing chances of photoinhibition in leaves should water availbility become limited, however with available water overall sensitivity to stress decreases. However, understory shaded/younger trees have comparitively less carbon reserve due to light limitation as a primary stressor [which differs among shade-tolerant/intolerant], decreased plant biomass, increased leaf biomass, and decreased evapotransition demands, with greater overall sensitivitty to stress. Larger carbon reserves in trees with greater light and water availability allows them to tolerate greater number of stressors compared to understory/younger trees that are able to tolerate only a fewer stressors making them overall more sensitive to stress [such as temperature]. Thus, overall stress sensitivity decreases with increasing height, light and age.[@niinemetsResponsesForestTrees2010a]
Their traits will be different from shade leaves of canopy trees as well. E.g.
They tend to have long-lived leaves with high LMA while shade leaves of canopy trees have large leaves with low LMA. Those differences in traits are likely to affect their thermal properties as well.

Passing light called sunflecks in the forest understory often caused by over-canopy vegetation gaps and wind speed cause a different type of morphological adaptation in shaded leaves [characterized by low LMA and greater effective leaf width] that make them efficient in capturing >50% of light through sunflecks for photosynthesis [@chazdonPhotosyntheticLightEnvironments1984]

[I think that this distinction should be addressed throughout the manuscript, so that it is always clear that the reader knows whether the cited references are about inter- or intra-specific variation, and that we address the different trait axes (Marielle made a comment on this earlier as well, that within species patterns may not follow leaf economics spectrum expectations). Ideally we would address the different categories, but if lack of information forces us to focus on just within species or just across species, this should still be made clear. photosynthesis, and canopy-level carbon uptake. They don't have to be treated separately, but it should always be clear which is being referred to. And in the consequences for scaling section this is of course also relevant.]

Increasing number of studies point towards the importance of incorporating vertical canopy leaf trait variation in modelling for leaf economic spectrum (LES) [@chenLeafEconomicsSpectrum2020a], leaf thermoregulation and thermal sensitivity which will provide a more accurate representation of within-canopy gradient response to increasing temperatures.[@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016; @niinemetsWorldwideAnalysisWithincanopy2015, @chenLeafEconomicsSpectrum2020, REFS)

[Add something about how leaf traits along the vertical canopy profile are different than the expectation from leaf economic spectrum (LES)? Chen et al. 2020 is a good paper to help frame this . They make an important point in that paper that we may also want to make here: that the LES was 'built' using mainly sun (upper canopy) leaves; given the vastly different light environments at the bottom of the canopy (and that most leaves in the canopy are likely to be shade leaves) there is a strong research need to better characterise trait relationships and leaf responses through the whole vertical canopy - hence the importance of this review. Keenan & Niinemets 2016  will also be helpful in making this point.]--M N. Smith

@@

```{r eval=FALSE, echo=FALSE}
# refs in this table need to go in the nocite field below (easiest in alphabetical order)
```
---
nocite: | 
  @mau_temperate_2018; @hernandezSimilarTemperatureDependence2020; @cobleHowVerticalPatterns2016a; @slotPhotosyntheticHeatTolerance2019; @valladaresShadeToleranceKey2008b; @beaumontVerticalGradientsLeaf2009; @sackHowStrongIntracanopy2006,@cobleLightDrivesVertical2014; @scartazzaInvestigatingEuropeanBeech2016; @martinCovarianceSunShade2020; @bolstad_foliar_1999; @harleyEffectsLightTemperature1996; @kullDistributionLeafPhotosynthetic1998; @kafutiFoliarWoodTraits2020; @duursmaVerticalCanopyGradients2006; @cobleLightDrivesVertical2014a; @vanwittenbergheVariabilityStomatalConductance2012; @poorterLeafOpticalProperties1995; @poorterMetaanalysisPlantResponses2019b; @cavaleri_foliar_2008; @niinemetsAcclimationHighIrradiance1998; Panditharathna et al. 2008; Baltzer and Thomas 2005; Kusi and Karasi 2020
---

>>>>

Forests are vertically and horizontally stratified, with trees of different sizes through various successional stages. Overstory trees form canopies that play a crucial role in moderating macro-microclimate [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017] by buffering extreme temperatures along with other macroclimatic conditions in the understory. This creates a vertical stratification of biophysical environments such as temperature, light, wind, humidity and Co2 which influences leaf temperature and metabolism along the gradient, and in turn whole plant performance [Bertrand et al. 2011].

@@

*(paragraph on leaf area distribution, including figure ([separate from or combined with current Fig. 2- see GitHub issue #15](https://github.com/EcoClimLab/vertical-thermal-review/issues/15))*
The vertical structure and hence, microenvironments of forest canopies vary seasonally (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019) and following disturbance [e.g., @parkerCanopyLightTransmittance2002 ; @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020].
In this review, we focus primarily on growing season conditions, including leaf phenology at the edges of the growing season but excluding periods (in deciduous forests) when no leaves are present.

Light conditions, specifically photosynthetically active radiation (PAR) varies along the vertical gradient with leaf area index (LAI), canopy height, canopy structure, across species and forest types [@koike_leaf_2001]. Generally, upper canopies in dense forests are exposed to greater PAR values and limit light to canopy interiors, lower canopy layers and understories. Shaded regions under canopy cover experience lower values and quality of PAR [in ranges of blue, orange and red]. The gradient of light is more pronounced under broadleaved forests than conifers[@lowmanForestCanopies1995; @aussenacInteractionsForestStands2000a; @niinemetsPackingPhotosyntheticMachinery2009; @poorterMetaanalysisPlantResponses2019b]
(Fig. `r fig_NEON_vertical`).

In heterogeneous canopies [with high gap fractions], distance from the outer canopy is a better proxy for light environment than height (Parker 1995) and lower canopy light environments are highly variable, ranging from high light gap to deeply shaded understory regions [@chazdonPhotosyntheticLightEnvironments1984; @tymenQuantifyingMicroenvironmentalVariation2017; @smithSeasonalDroughtrelatedChanges2019]

Tropical forests with dense overstory limit light to understories and subsequent layers, where tropical forest floors receive approximately 1-2%  of the light incident to the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], whereas mixed-wood boreal and Northern temperate forest floors receive around 4-5% of full sunlight [@bartemucciOverstoryInfluencesLight2006]. However, vertical canopy layers and forest floors also receive light permeating through canopy gaps and movements caused by wind speed called sunflecks that are often intermittent and breif. Sunflecks can provide 10-80% of photon flux density for photosynthesis in canopy layers and understory depending on canopy attributes such as foliage density, height and architecture [Way and Pearcy 2012, REF]

This diffence in light regime along a vertical profile is an important driver of plasticity in leaf traits, leaf physiology, adaptation, and thermal sensitivity along the gradient. Leaves along a vertical intra-canopy gradient, as well as in inter-canopies, thus exhibit a variation in leaf trait adapaptions to a range of high light to shaded-ness [@michaletzEnergeticCarbonEconomic2016; @fauset_differences_2018; @niinemetsResponsesForestTrees2010, @sackHowStrongIntracanopy2006; @doughtyAreTropicalForests2008; @niinemetsControversyTraitsConferring2006].

@@

Wind speed in the upper canopy of a semi-deciduous tropical forest has been observed to be higher in dry season than in upper canopy wet season [@rey-sanchez_spatial_2016] [*double check*]. 
Greater upper canopy wind speeds result in higher boundary layer conductance ($g_{a}$) for canopy leaves, which increases linearly with wind velocity [@daudetWindSpeedLeaf1999a]. Large $g_{a}$ to heat transfer results in smaller difference in $T_{air}$ and $T_{leaf}$, this is more typical of small leaves in upper canopy that make them more advantageous in dissapating heat efficiently than larger leaves[@bauerleSeparatingFoliarPhysiology2011]. In low wind conditions, because $g_{a}$ to heat transfer is small enough, $T_{leaf}$ in upper canopy can become substantially elevated than $T_{air}$ under high radiation [@martin_boundary_1999]. Additionally, thickness of air layer is proportional to boundary layer resistance at the surface of the leaf through which water vapor diffuses after leaving the stomata [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a; @stokesBoundaryLayerConductance2006] this tends to be greater in the understory than canopy? [ref]

@@

Seasonally, maximum air temperatures were higher above than below tropical forest canopies during wet seasons in Panama [@rey-sanchez_spatial_2016], temperate forests in Switzerland [@vonarxSpatiotemporalEffectsForest2012a] and coastal Brazil [@fauset_differences_2018, Tymen et al. 2017].Similarly, maximum temps were higher and humidity lower in gaps vs understory regions (both at 1.5 m above the ground, see Fig. 4 - again illustrating buffering effect of the canopy).However, similar maximum temperatures have been observed during the dry season of a semi-deciduous tropical forest in Panama partially because in the dry season some canopy trees were leafless, resulting in more light reaching the sub-canopy, and more air movement between subcanopy and above-canopy air[@rey-sanchez_spatial_2016], similar results were observed in in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate].

@@

RH, in combination with temperature, determines vapor pressure deficit, $VPD$. Atlantic forests during wet season tended to have higher $T_{air}$, and $VPD$ in upper canopy and canopy gaps than lower in understory where $T_{air}$ and $VPD$ were lower [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017].  
This means that canopy leaves tend to be exposed to higher *evaporative demand*, higher stomtal contrains and lower osmotic potential. [@niinemetsPhotosyntheticAcclimationSimultaneous2004a]. In North Eastern US temperate forests, upper canopy was observed to be more sensitive to $VPD$, whereas understory was insensitive [@rollinsonClimateSensitivityUnderstory, 2020]

**Finally, carbon dioxide (CO~2~) concentrations tend to be higher in the understory.** 
	Higher in understory, particularly at dusk [@koike_leaf_2001]. Higher in the understory overnight; difference persists during the day but is very small [@yang_spatial_1999]. Differences in concentration are by far most pronounced near ground level, as well, concentrations of (CO~2~) mole fraction values have been observed to be higher in lower canopy leaves compared with upper canopy leaves [@niinemetsPhotosyntheticAcclimationSimultaneous2004a]

@@

*Presumably, buffering would also be affected by canopy roughness, which affects turbulent air flow and the canopy boundary layer.*
Taller trees don't necessarily increase buffering [@zellweger_seasonal_2019]. "They found that taller forests (which coincided with forests of higher LAI) had lower mean and max values of VPD and temperature" [Jucker, T., Hardwick, S.R., Both, S., Elias, D.M., Ewers, R.M., Milodowski, D.T., Swinfield, T. and Coomes, D.A., 2018. Canopy structure and topography jointly constrain the microclimate of human-modified tropical landscapes. Global change biology, 24(11), pp.5243-5258]
.See Jucker et al. 2018 - max height was highly correlated to both LAI and surface roughness

Species that have higher shade casting ability such as braodleaved trees increase T buffering [@zellweger_seasonal_2019,@vonarxSpatiotemporalEffectsForest2012]
The strength of buffering also varies with respect to geographic and climatic factors.
* Distance to coast, topographic position, elevation [@zellweger_seasonal_2019]
* [@davis_microclimatic_2019]--Proximity to riparian areas tends to sustain buffering, protecting understory temperatures even during increasing temperatures [@davis_microclimatic_2019]
However vertical climate gradients within forests are known to be steeper than those driven by elevation and latitude [@nakamuraForestsTheirCanopies2017; @martinCovarianceSunShade2020a]

@@

Leaf traits are shaped on the basis of light and height in forest vertical strata, and show different degrees of foliar plasticity in accordance with the intensity and availability of light. After the formation of leaves to integrated light, it takes 30-60 days for traits to stabalize in woody species [@niinemetsLeafAgeDependent2016a]. A network of phytohormones in leaves perceive the intensity of light falling on the surface and thus respond through receptors that enable acclimation or avoidance depending on the light scenario [Casal, 2013]. As a result, degrees of high irradiance gives rise to variations of sun-leaf morphology compared to degrees of low irradiance-- characterized by shade-- shapes variations of shade-leaf morphology [@poorterMetaanalysisPlantResponses2019b; Sack et al 2006], which ultimately also shapes leaf structural, chemical and physiological traits as well [Keenan and Niinemets et al. 2016]. Increasing number of studies point towards the importance for incorporating sun and shade in modelling leaf economic spectrum (LES) within-canopy gradient[@chenLeafEconomicsSpectrum2020a] because LES was 'built' using mainly sun (upper canopy) leaves; given the vastly different light environments at the bottom of the canopy (and that most leaves in the canopy are likely to be shade leaves) there is a strong research need to better characterise trait relationships and leaf responses through the whole vertical canopy [Keenan and Niinemets 2016]. However, there is also concern for ambiguity in simplified cateogorization of 'sun' and 'shade' leaves due to variation based on intensities of light within-canopy, across functional groups, in canopy gaps, height and ontogeny [@ishiiVariationIntracrownLeaf2018; @niinemetsWorldwideAnalysisWithincanopy2015] as well as geographically [@ishiiVariationIntracrownLeaf2018].

**Inter-species vertical gradient leaf trait variation**

Taller trees in dense forests tend to be exposed to greater irradiance than lower canopy and shorter understory trees shaded by overstory. As a result the combination of stressors taller trees with foliage exposed to greater PAR and $VPD$ experience are different from the combination of stressors experienced by understory shorter trees with shaded/semi-shaded leaves [@niinemetsResponsesForestTrees2010a] 
"Spending their lifecycle in the shade means they have different pressures [such as competition for light or adapting leaf traits to maximize light capture and use efficiency] than plants that get the bulk of their carbon from sun-exposed leaves, but can maintain a few layers of shade leaves as well."[-Martijn]

Cateogorization of sun and shade tolerance from @niinemetsControversyTraitsConferring2006 are as follows: "The five-level scale used for shade tolerance (1,very intolerant; 2, intolerant; 3, moderately tolerant; 4,tolerant; 5, very tolerant) corresponds approximately to the following light availabilities expressed as percentage of full sunlight: 1,>50%; 2, 25-50%; 3, 10-25%; 4, 5-10%; 5, 2-5%".

"Additionally, shade leaves in the understory are of three categories; 1. shade leaves of canopy trees--top of the tree is sun exposed, bottom is shaded. 2. seedlings/saplings of canopy trees that will eventually, if they survive, reach the canopy. 3 understory plants that complete their lifecycle in the shaded understory. These groups are likely to have different traits, and if we expect'shade' plants to have a different sensitivity to climate change than sun leaves, the consequences of climate change for demography and species community composition of the forest will depend strongly on the category of shade plants being considered".[refs, -Martijn Slot]

Leaf traits have been observed to exhibit greater plasticity on the basis of area than mass 
within-canopy and across inter-canopies [Keenan and Niinemetes, 2016]

**Inter-species ontogeny and height leaf trait variation**

@@

Additionally, increased exposure to light increases carbon reserve, plant height, foliage robustness,  while also increasing evapotranspiration, decreasing hydraulic conductivity, increasing chances of photoinhibition in leaves should water availbility become limited, however with available water overall sensitivity to stress decreases. Understory shaded/younger trees have comparitively less carbon reserve due to light limitation as a primary stressor [which differs among shade-tolerant/intolerant], decreased plant biomass, increased leaf biomass, and decreased evapotransition demands, with greater overall sensitivitty to stress. Larger carbon reserves in trees with greater light and water availability allows them to tolerate greater number of stressors compared to understory/younger trees that are able to tolerate only a fewer stressors making them overall more sensitive to stress [such as temperature]. Thus, overall stress sensitivity decreases with increasing height, light and age.[@niinemetsResponsesForestTrees2010a]
Their traits will be different from shade leaves of canopy trees as well. E.g.
They tend to have long-lived leaves with high LMA while shade leaves of canopy trees have large leaves with low LMA. Those differences in traits are likely to affect their thermal properties as well.

**Intra-canopy on the same species leaf trait variation**

Several studies point towards common variations and differences in leaf trait plasticity within-species and at an intra-canopy gradient which shapes leaf structural and biochemical traits [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020]. The differences among these traits contribute to leaf heat dissapation strategies and leaf temperature across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table 1. summaries relevant traits for thermal sensitivity.

[I think that this distinction should be addressed throughout the manuscript, so that it is always clear that the reader knows whether the cited references are about inter- or intra-specific variation, and that we address the different trait axes (Marielle made a comment on this earlier as well, that within species patterns may not follow leaf economics spectrum expectations). Ideally we would address the different categories, but if lack of information forces us to focus on just within species or just across species, this should still be made clear. photosynthesis, and canopy-level carbon uptake. They don't have to be treated separately, but it should always be clear which is being referred to. And in the consequences for scaling section this is of course also relevant.]
[Add something about how leaf traits along the vertical canopy profile are different than the expectation from leaf economic spectrum (LES)? Chen et al. 2020 is a good paper to help frame this . They make an important point in that paper that we may also want to make here: that the LES was 'built' using mainly sun (upper canopy) leaves; given the vastly different light environments at the bottom of the canopy (and that most leaves in the canopy are likely to be shade leaves) there is a strong research need to better characterise trait relationships and leaf responses through the whole vertical canopy - hence the importance of this review. Keenan & Niinemets 2016  will also be helpful in making this point.]--M N. Smith

@@

```{r eval=FALSE, echo=FALSE}
# refs in this table need to go in the nocite field below (easiest in alphabetical order)
```
---
nocite: | 
  @mau_temperate_2018; @hernandezSimilarTemperatureDependence2020; @cobleHowVerticalPatterns2016a; @slotPhotosyntheticHeatTolerance2019; @valladaresShadeToleranceKey2008b; @beaumontVerticalGradientsLeaf2009; @sackHowStrongIntracanopy2006,@cobleLightDrivesVertical2014; @scartazzaInvestigatingEuropeanBeech2016; @martinCovarianceSunShade2020; @bolstad_foliar_1999; @harleyEffectsLightTemperature1996; @kullDistributionLeafPhotosynthetic1998; @kafutiFoliarWoodTraits2020; @duursmaVerticalCanopyGradients2006; @cobleLightDrivesVertical2014a; @vanwittenbergheVariabilityStomatalConductance2012; @poorterLeafOpticalProperties1995; @poorterMetaanalysisPlantResponses2019b; @cavaleri_foliar_2008; @niinemetsAcclimationHighIrradiance1998; @panditharathnaChangeLeafStructure2008; @baltzerLeafOpticalResponses2005; @kusiPlasticLeafMorphology2020
---
